Characterizing subarctic European vegetation for land surface modeling of ozone pollution and climate risk

### 2021-09-28 VegOzone.tex
Restructure text

In summary, 2018 had a significant number of days that were warmer, drier, and brighter than the climatological average, while 2019 was a rather average year. Ozone was not significantly enhanced during the growing season in 2018, but a substantially higher number of peak \chem{[O_3]} were observed in 2018 than in 2019.

\section{DO3SE modeling}
\label{sec:do3se}
In the following, we investigate the difference between 2018 and 2019 concerning stomatal ozone uptake and assess which of the environmental factors might be most influential in determining \chem{O_3} damage risk. We model ozone uptake of natural and semi-natural vegetation using the $\mathrm{DO_3SE}$ model and develop bespoke subarctic parameterizations for common plant species. First, we give an account of the methodology for deriving bespoke parameters, after which we present and discuss the modeling results. The relevant parts of $\mathrm{DO_3SE}$ model are summarized in Appendix~\ref{appendix:do3se_model}.

@@

We deduced the dominant species on-site at the NIBIO Environment Centre Svanhovd and from Fig.~\ref{fig:svanhovd_research_station} and found perennial grassland, birch (generalized as deciduous trees), and Scots pine (generalized as coniferous trees). Default parameters for these species or PFTs are derived from \citet{EP:Simpson2007,GCB:Mills2011,ICP:MappingManual2017} and will be referred to as \emph{mapping manual} (MM) parameterizations in the following. We base our parameterization of coniferous trees on the MM's Boreal Norway spruce and deciduous trees on the MM's silver birch. For a comprehensive list of model parameters, consult Supplement~Table~S1.

Initial test simulations with the MM PFTs revealed an unrealistically low stomatal conductance at leaf-level ($G_\mathrm{sto}^\mathrm{leaf}$) in 2019 particularly for perennial grassland (Appendix~Fig.~\ref{fig:pody_mm_composit}). Substantial stomatal conductance occurred only during an extended warm period in late July. In ecological terms, this means that there was almost no growth of grass in the summer of 2019 -- a prediction which is easily falsified by reality. We identified $f_\mathrm{T}$ as abnormally low, being the limiting factor of stomatal conductance in perennial grassland. Therefore, we saw the necessity to propose bespoke parameterizations describing an acclimation of of the PFTs to the climatic conditions in the target region. Note, however, that these parameterizations are hypothetical and have yet to be verified by experiments.

We assume that perennial vegetation will likely maximize carbon acquisition during the short subarctic growing season. Because the mapping manual version of the $\mathrm{DO_3SE}$ model currently does not simulate net photosynthesis ($A_\mathrm{n}$), we assume a first order proportionality between $A_\mathrm{n}$ and $g_\mathrm{sto}$ \citep{GCB:Medlyn2011} and effectively tune the temperature and light response functions ($f_\mathrm{T}$ and $f_\mathrm{light}$) for higher $\left<g_\mathrm{sto}\right>/g_\mathrm{max}$ (see Appendix~\ref{appendix:do3se_model} for details) to acclimate the growth potential to the local climate. We further assume that $f_\mathrm{VPD}$ and $f_\mathrm{SWP}$ suit our vegetation types. We aim to increase the overlapping area between the probability density function (PDF) and the respective response function.

We define two different hypothetical temperature acclimations: \emph{subarctic} and \emph{cold}. For either, we assume that the hypothetical species are more tolerant to cold conditions and shift $f_\mathrm{T}$ towards colder temperatures accordingly. We construct \emph{cold} as representative for a species that is more tolerant to cold temperatures, but slightly less efficient at warm temperatures compared to MM. In the same way, \emph{subarctic} refers to a hypothetical species which is very tolerant to cold but is sensitive to high temperatures, and most efficient at cool temperatures ($T_\mathrm{opt}$ close to the climatological mean temperature of 1992--2000).
We show two climatological reference periods 1990s and 2000s to indicate that the vegetation is subject to ongoing climate change and hence continuous acclimation. The most extreme of the two acclimation types is \emph{subarctic}. For perennial grassland the resulting response functions are depicted together with a five month growing season temperature PDF of the two periods 1992-2000 and 2011-2019 in Fig.~\ref{fig:f_temp_grassland}a.

The resulting response functions for coniferous and deciduous trees are displayed in Appendix~Figs.~\ref{fig:f_temp_spruce} and \ref{fig:f_temp_birch}, respectively.

Norway spruce was experimentally found to be active already at rather low air temperatures and can reach $60\,\unit{\%}$ photosynthetic activity as early as \unit{doy}~100 \citep{TB:Kolari2007, TP:Wallin2013}. Based on the time series of \chem{CO_2} uptake and temperature at observation sites in southern ($\rightarrow$ \emph{cold}) and northern ($\rightarrow$ \emph{subarctic}) Finland \citep{TB:Kolari2007}, we estimate the optimal temperature interval $10\,\unit{^\circ C}\le T_\mathrm{opt} \le 15\,\unit{^\circ C}$. We assume similar temperature acclimations for deciduous trees. All bespoke temperature parameters are listed in Table~\ref{tab:sensitivity_tests_temp}.

With respect to light conditions, we assume that the light threshold for a $50\,\unit{\%}$ opening of plants' stomata may differ in subarctic species compared to species in less extreme climates. Analytically, we derive the inverse function $f_{\mathrm{light},k}^{-1}$ of Eq.~(\ref{eq:flight}) for each PFT or species $k$
\begin{equation}
  \gamma_k(f_\mathrm{light}) := f_{\mathrm{light}, k}^{-1} = -\frac{\ln(1-f_\mathrm{light})}{\alpha_k}.
  \label{eq:inverse_function}
\end{equation}
First, we solve Eq.~(\ref{eq:inverse_function}) for the MM default value $\alpha_\mathrm{MM}^k$ for each PFT or species $k$ at $0.5$ 
\begin{equation}
  \gamma^k(0.5) = -\frac{\ln(0.5)}{\alpha_\mathrm{MM}^k},
  \label{eq:inverse_function_halfway}
\end{equation}
and define a variation $\gamma' = \gamma \cdot \eta$ with $\eta \in \{0.8, 1.2\}$. We invert Eq.~(\ref{eq:inverse_function}) for $\alpha$ and find $\alpha'(\gamma')$. The resulting functions for perennial grassland are shown in Fig.~\ref{fig:f_temp_grassland}b) (for deciduous and coniferous trees refer Appendix~Figs.~\ref{fig:f_temp_spruce}b)--\ref{fig:f_temp_birch}b)). All derived parameters are tabulated in Table~\ref{tab:sensitivity_tests_light}.

@@

As indicated above, a better acclimation to prevailing climate conditions will result in a higher average $\left<g_\mathrm{sto}\right>/g_\mathrm{max}$. To compare the bespoke parameterizations, we propose the following metric. We calculate an average relative stomatal conductance $\left<g_\mathrm{sto}\right>/g_\mathrm{max}$ (Eq.~(\ref{eq:stomatal}) with $f_\mathrm{SWP}=1$) at noon ($11\,\unit{am}-1\,\unit{pm}$ local time) and in the morning ($5-9\,\unit{am}$ local time) for the whole record of meteorological data and apply mean and standard deviation. We presume that $\left<g_\mathrm{sto}\right>/g_\mathrm{max}$ around noon (highest light intensity) is a good proxy for \chem{CO_2} uptake efficiency (growth) which should be maximized. In additional, a small deviation between noon and morning and a low standard deviation would indicate a higher robustness to variability in growing conditions. From the results shown in Fig.~\ref{fig:javis_func_opt_mean}, we identify \emph{subarctic}, PPFD0.8 as the best acclimation for all PFTs, based on the criteria. As expected due to the small adjustments, coniferous trees display the lowest deviation between the different types of acclimation, while the divergence for perennial grassland is substantial as a response to the stronger proposed temperature acclimation.

>>>>

In summary, 2018 had a significant number of days that were warmer, drier, and brighter than the climatological average, while 2019 was a rather average year. Ozone was not significantly enhanced during the GS 2018/19, but a substantially higher number of peak \chem{[O_3]} were observed in 2018 than in 2019.

\section{DO3SE modeling}
\label{sec:do3se}
In the following, we investigate the difference between 2018 and 2019 concerning stomatal ozone uptake and assess which of the environmental factors might be most influential in determining \chem{O_3} damage risk. We model ozone uptake of natural and semi-natural vegetation using the $\mathrm{DO_3SE}$ model \citep{ACP:Bueker2012} and develop bespoke parameterizations for common plant species in our focus area. First, we give an account of the methodology for deriving bespoke parameters, after which we present and discuss the modeling results. The relevant equations of the $\mathrm{DO_3SE}$ model are summarized in Appendix~\ref{appendix:do3se_model}.

@@

We deduced the dominant species on-site at the NIBIO Environment Centre Svanhovd and from Fig.~\ref{fig:svanhovd_research_station} and found perennial grassland, birch (generalized as deciduous trees), and Scots pine (generalized as coniferous trees). Default parameters for these species or PFTs are derived from \citet{EP:Simpson2007,GCB:Mills2011,ICP:MappingManual2017} and will be referred to as mapping manual (MM) parameterizations in the following. We base our parameterization of coniferous trees on the MM's Boreal Norway spruce, deciduous trees on the MM's silver birch, and perennial grassland on MM's perennial grassland for central Europe. For a comprehensive list of model default parameters, consult Supplement~Table~S1.

Initial test simulations with the MM PFTs revealed an unrealistically low stomatal conductance at leaf-level ($G_\mathrm{sto}^\mathrm{leaf}$) in 2019 particularly for perennial grassland (Appendix~Fig.~\ref{fig:pody_mm_composit}). Substantial stomatal conductance occurred only during an extended warm period in late July. In ecological terms, this means that there was almost no growth of grass in the summer of 2019 -- a prediction which is easily falsified by reality. We identified $f_\mathrm{T}$ as abnormally low, being the limiting factor of stomatal conductance in perennial grassland. Therefore, we saw the necessity to propose bespoke parameterizations describing an acclimation of the default PFTs to the climatic conditions in the focus area. Note, however, that these parameterizations are hypothetical and have yet to be verified by experiments.

We assume that perennial vegetation will likely maximize carbon acquisition during the short subarctic growing season. Because the mapping manual version of the $\mathrm{DO_3SE}$ model currently does not simulate net photosynthesis ($A_\mathrm{n}$), we assume a first order proportionality between $A_\mathrm{n}$ and $g_\mathrm{sto}$ \citep{GCB:Medlyn2011} and effectively tune the temperature and light response functions ($f_\mathrm{T}$ and $f_\mathrm{light}$) for higher $\left<g_\mathrm{sto}\right>/g_\mathrm{max}$ (see Appendix~\ref{appendix:do3se_model} for details) to acclimate the growth potential to the local climate. Further, we presume that $f_\mathrm{VPD}$ and $f_\mathrm{SWP}$ suit our vegetation types and no acclimation is necessary for these.

We define two different hypothetical temperature acclimations: subarctic and cold. We construct \emph{cold} as representative for a species that is more tolerant to cold temperatures, but slightly less efficient at warm temperatures compared to MM. This is accomplished by moving $T_\mathrm{opt}$ towards cooler temperatures while keeping the other parameters fixed to default values. In the same way, \emph{subarctic} constructed to represent a species which is very tolerant to cold but sensitive to high temperatures, and most efficient at cool temperatures. Therefore, we shift $T\mathrm{min}$ and $T\mathrm{max}$ to colder temperatures and choose $T_\mathrm{opt}$ close to the climatological mean temperature of 1992--2000. In other words, we aim to increase the overlapping area between the $f_\mathrm{temp}$ and the temperature PDF of the GS. The most extreme of the two acclimation types is \emph{subarctic}.
In Fig.~\ref{fig:f_temp_grassland}a, we show the resulting $f_\mathrm{temp}$ for perennial grassland together with temperature PDFs deduced from five months GS in two climatological periods. The periods 1990s and 2000s have been chosen to indicate that the vegetation is subject to ongoing climate change Thus, the predominant acclimation at present is not known.

The resulting response functions for coniferous and deciduous trees are displayed in Appendix~Figs.~\ref{fig:f_temp_spruce}a and \ref{fig:f_temp_birch}a, respectively. We base our temperature acclimation of coniferous trees on experimental results on Norway spruce which were found to be active already at rather low air temperatures and can reach $60\,\unit{\%}$ photosynthetic activity as early as \unit{doy}~100 \citep{TB:Kolari2007, TP:Wallin2013}. In particular, we use the time series of \chem{CO_2} uptake and temperature at observation sites in southern ($\rightarrow$ \emph{cold}) and northern ($\rightarrow$ \emph{subarctic}) Finland by \citet{TB:Kolari2007} to estimate the optimal temperature interval ($10\,\unit{^\circ C}\le T_\mathrm{opt} \le 15\,\unit{^\circ C}$). We presume a similar temperature acclimation for deciduous trees. All bespoke temperature parameters are listed in Table~\ref{tab:sensitivity_tests_temp}.\\

Regarding the opening extent of stomata at low light conditions, we look at the PPFD threshold for a $50\,\unit{\%}$ opening of plants' stomata. We assume that this threshold may differ in subarctic species compared to species in less extreme climates. We derive the inverse function $f_{\mathrm{light},k}^{-1}$ of Eq.~(\ref{eq:flight}) analytically for each PFT or species $k$
\begin{equation}
  \gamma_k(f_\mathrm{light}) := f_{\mathrm{light}, k}^{-1} = -\frac{\ln(1-f_\mathrm{light})}{\alpha_k}.
  \label{eq:inverse_function}
\end{equation}
First, we solve Eq.~(\ref{eq:inverse_function}) for the MM default value $\alpha_\mathrm{MM}^k$ at $0.5$ 
\begin{equation}
  \gamma^k(0.5) = -\frac{\ln(0.5)}{\alpha_\mathrm{MM}^k},
  \label{eq:inverse_function_halfway}
\end{equation}
and define a variation $\gamma' = \gamma \cdot \eta$ with $\eta \in \{0.8, 1.2\}$ and invert Eq.~(\ref{eq:inverse_function}) for $\alpha$ to find $\alpha'(\gamma')$. The resulting functions for perennial grassland (PPFD0.8, PPFD1.2) are shown in Fig.~\ref{fig:f_temp_grassland}b) (for deciduous and coniferous trees refer Appendix~Figs.~\ref{fig:f_temp_spruce}b and \ref{fig:f_temp_birch}b). All derived parameters are tabulated in Table~\ref{tab:sensitivity_tests_light}.

@@

As indicated above, a better acclimation to prevailing climate conditions is characterized by a higher average $\left<g_\mathrm{sto}\right>/g_\mathrm{max}$. To compare our bespoke parameterizations, we propose the following metric. We calculate an average relative stomatal conductance $\left<g_\mathrm{sto}\right>/g_\mathrm{max}$ (Eq.~(\ref{eq:stomatal}) with $f_\mathrm{SWP}=1$) at noon ($11\,\unit{am}-1\,\unit{pm}$ local time) and in the morning ($5-9\,\unit{am}$ local time) for the whole record of meteorological data and apply mean and standard deviation. We presume that $\left<g_\mathrm{sto}\right>/g_\mathrm{max}$ around noon (highest light intensity) is a good proxy for \chem{CO_2} uptake efficiency (growth). In additional, a small deviation between noon and morning and a low standard deviation would indicate a higher robustness to variability in growing conditions. Based on the criteria, we identify \emph{subarctic}--PPFD0.8 as the best parameterization for all PFTs (Fig.~\ref{fig:javis_func_opt_mean}). As expected due to the small adjustments, coniferous trees display the smallest differences between the different types of acclimation, while the differences for perennial grassland are substantial as a response to the stronger proposed temperature acclimation.
